---
title: "PROPHECY Consensus Variants"
author: "Stevie Pederson"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
editor_options: 
  chunk_output_type: console
---

```{r setup, incude = FALSE}
knitr::opts_chunk$set(
  message = FALSE, warning = FALSE, fig.height = 8, fig.width = 10
)
```

```{r packages}
library(transmogR)
library(VariantAnnotation)
library(rtracklayer)
library(readr)
library(plyranges)
library(scales)
library(pander)
panderOptions("big.mark", ",")
```


## Introduction

This documents the creation of the PROPHECY consensus variant set and the accompanying VCF.
The original VCF was provided by Dr Yassine Souilmi as `prophecy_unrelated_PASS_MAF0.5_sites-only_merged.vcf.gz` containing sites passing QC which were detected in >50% of unrelated individuals, using >3 familial steps to consider two individuals unrelated.


## Variant Preparation

```{r load-vcf}
## Load the VCF
vcf <- VcfFile(
	here::here(
	  "variants", "prophecy_unrelated_PASS_MAF0.5_sites-only_merged.vcf.gz"
	)
)
sq <- vcf %>% 
  seqinfo() %>% 
  keepStandardChromosomes() %>% 
  dropSeqlevels("chrM") 
genome(sq) <- "GRCh38"
isCircular(sq) <- rep(FALSE, length(sq))
gr <- GRanges(sq)
param <- ScanVcfParam(fixed = "ALT", info = NA, which = gr)
vcf_ranges <- readVcf(vcf, param = param) |> rowRanges()
n <- length(vcf_ranges)
vcf_ranges <- vcf_ranges[!duplicated(vcf_ranges)]
seqlevels(vcf_ranges) <- seqlevels(sq)
seqinfo(vcf_ranges) <- sq
```

After removing any duplicated variants, the complete set of unfiltered variants contained `r comma(length(vcf_ranges))` loci.

### Multi-Allelic and Missing Sites

```{r check-multi-allelic}
alts <- vcf_ranges$ALT
names(alts) <- names(vcf_ranges)
alts <- unlist(alts)
multi <- duplicated(names(alts))
alts <- alts[!multi]
vcf_ranges$ALT <- unname(alts)
```

`r comma(sum(multi))` multi-allelic sites were then identified and only the first alternate allele was retained

```{r w0}
w0 <- width(vcf_ranges$ALT) == 0
vcf_ranges <- vcf_ranges[!w0] 
```

An additional `r comma(sum(w0))` sites were removed as having no alternate allele provided by GATK

### Overlapping Variants

```{r tidy-substitutions}
subs <- (nchar(vcf_ranges$REF) == nchar(vcf_ranges$ALT)) & nchar(vcf_ranges$REF) > 1
subs_to_snp <- vcf_ranges[subs] %>%
	GPos() %>%
	mutate(
		REF = vcf_ranges$REF[subs] %>%
			unlist() %>% 
			as.character() %>%
			strsplit(split = "") %>%
			.[[1]],
		ALT = vcf_ranges$ALT[subs] %>%
			unlist() %>% 
			as.character() %>%
			strsplit(split = "") %>%
			.[[1]]
	) %>%
	subset(REF != ALT) %>%
	as("GRanges") %>%
	setNames(paste0(as.character(.), ":", .$REF, ":", .$ALT)) %>%
	mutate(REF = DNAStringSet(REF), ALT = DNAStringSet(ALT))
vcf_ranges <- vcf_ranges[!subs] %>%
	select(REF, ALT) %>%
	c(subs_to_snp) %>%
	sort() 
```

Next, `r comma(sum(subs))` sites were identified which were effectively substitutions but extending their range beyond being SNPs. 
These were expanded into a series of SNPs and the sites retained were those for which `REF` and `ALT` alleles were different.

### Incorrectly Coded InDels

```{r tidy-insertions}
bad_insertion <- width(vcf_ranges$ALT) > width(vcf_ranges) & width(vcf_ranges) > 1
alts <- vcf_ranges[bad_insertion]$ALT
refs <- vcf_ranges[bad_insertion]$REF
new_alts <- Views(
	unlist(alts), 
	start = c(1, 1+cumsum(width(alts))[-length(alts)]), 
	width = width(alts)
) %>%
	resize(
		width = width(alts) - width(refs) + 1, fix = "end"
	) %>% 
	DNAStringSet()
new_refs <- Views(
	unlist(refs), 
	start = c(1, 1+cumsum(width(refs))[-length(refs)]), 
	width = width(refs)
) %>%
	resize(width = 1, fix = 'end') %>%
	DNAStringSet()
vcf_ranges[bad_insertion] <- resize(
	vcf_ranges[bad_insertion], width = 1, fix = 'end'
)
vcf_ranges[bad_insertion]$REF <- new_refs
vcf_ranges[bad_insertion]$ALT <- new_alts
```

Insertions are generally defined as replacing a 1nt sequence with a >1nt sequence.
The `r comma(sum(bad_insertion))` loci not conforming to this rule were then identified and restructured appropriately.


```{r tidy-deletions}
bad_deletion <- width(vcf_ranges$ALT) < width(vcf_ranges) & width(vcf_ranges$ALT) > 1
alts <- vcf_ranges[bad_deletion]$ALT
refs <- vcf_ranges[bad_deletion]$REF
new_alts <- Views(
	unlist(alts), 
	start = c(1, 1+cumsum(width(alts))[-length(alts)]), 
	width = width(alts)
) %>%
	resize(width = 1, fix = "end") %>% 
	DNAStringSet()
new_refs <- Views(
	unlist(refs), 
	start = c(1, 1+cumsum(width(refs))[-length(refs)]), 
	width = width(refs)
) %>%
	resize(
		width = width(refs) - width(alts) + 1, fix = "end"
	) %>% 
	DNAStringSet()
vcf_ranges[bad_deletion] <- resize(
	vcf_ranges[bad_deletion], width = width(new_refs), fix = 'end'
)
vcf_ranges[bad_deletion]$REF <- new_refs
vcf_ranges[bad_deletion]$ALT <- new_alts
```

Similarly, deletions are generally defined as the replacement of >1nt sequences with a single nucleotide, and the `r comma(sum(bad_deletion))` sites which did not conform to this structure were identified and restructured.

### Final Corrections

```{r}
too_long <- width(vcf_ranges) > 50 | width(vcf_ranges$ALT) > 50
dups <- duplicated(vcf_ranges)
vcf_ranges <- vcf_ranges[!too_long & !dups] %>% 
  sort() %>% 
	mutate(
		ALT = as(ALT, "DNAStringSetList")
	)
```

The complete set of `r comma(length(vcf_ranges))` variants was then filtered a final time to remove the `r comma(sum(dups))` duplicate entries introduced by the above restructuring, and the remaining `r sum(too_long)` InDels which extended beyond 50bp.
This reduced the set of `r comma(length(dups))` to the final set of `r comma(length(vcf_ranges))` fitting into the following categories:

`r pander(table(varTypes(vcf_ranges)))`


## Data Export

```{r export}
new_vcf <- here::here("variants", "PROPHECY_unrel_edited_0.5.vcf")
writeVcf(
	VCF(granges(vcf_ranges), fixed = mcols(vcf_ranges)),
	new_vcf
)
write_rds(
  vcf_ranges, here::here("variants", "PROPHECY_unrel_edited_0.5.rds"),
  compress = "gz"
)
```

